---
title: "**Advanced Data Programming with R - Assignment 1**"
author: "Vanessa Carvalho - 25267189"
date: today
date-format: long
format: 
  revealjs: 
    theme: default
    css: assignment_ctheme.css 
    slide-number: true
    toc: false
    center-title-slide: true
    transition: slide
    code-fold: true
    code-overflow: wrap
    fontsize: 1.2em
editor: visual
execute: 
  eval: true
  echo: true
---
# Introduction

For this first Assignment, I have selected [Brazil]{style="color: green;"}, [Angola]{style="color: red;"} and the [United Kingdom]{style="color: blue;"} to explore and analyse their *Economic, Social, Environmental, Health, Education, Development and Energy* data.  

As countries with quite a different development throughout the decades, I have selected the following specific indicators to be used in **comparisons** between each of them:

1. Life expectancy at birth (total) years
2. Net migration
3. Unemployment, total (% of total labor force) (modeled ILO estimate)

# Description of datasets and indicators

::::: columns 
::: {.column width="30%"}

*Life expectancy*

Indicates the number of **years** a newborn infant would live if prevailing patterns of mortality at the time of its birth were to stay the same throughout its life.

:::
::: {.column width="30%"}
*Net migration*

The net **total of migrants** during the period, that is, the number of immigrants minus the number of emigrants, including both citizens and non citizens.

:::
::: {.column width="30%"}
*Unemployment*

The share of the labor force that is **without work** but available for and seeking employment.

:::
:::::

\newpage

## Data Import and Overview

Here we import the datasets on Education, Health, and Energy using `fread()` from the `data.table` package.

```{r} 
#| echo: true
#| results: hide 
library(data.table)

#read each dataset
bra <- fread("indicators_bra.csv")
ago <- fread("indicators_ago.csv")
gbr <- fread("indicators_gbr.csv")

```

---

## Resumed Version of the Dataset (Table Preview)

Each table includes the first 6 rows of the datasets

### - [Brazil]{style="color: green;"} demographics

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 2
library(knitr)

# this code is for printing only the first 5 rows, minus the second one
bra_preview <- head(bra, 6)
bra_preview <- bra_preview[-2]
kable(bra_preview)

```

---

### - [Angola]{style="color: red;"} demographics

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 2
library(knitr)

# this code is for printing only the first 5 rows, minus the second one
ago_preview <- head(ago, 6)
ago_preview <- ago_preview[-2]
kable(ago_preview)

```

---

### - [United Kingdom]{style="color: blue;"} demographics

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 2
library(knitr)

# this code is for printing only the first 5 rows, minus the second one
gbr_preview <- head(gbr, 6)
gbr_preview <- gbr_preview[-2]
kable(gbr_preview)

```
# Data Cleaning

In this step each dataset will be cleaned by *keeping only my target indicators* and **removing the duplicates**

```{r}
# Defining indicators to keep

indicators_keep <- c("Life expectancy at birth, total (years)", "Net migration", "Unemployment, total (% of total labor force) (modeled ILO estimate)")

# Filter the datasets
bra_clean <- bra[`Indicator Name`%in% indicators_keep]
ago_clean <- ago[`Indicator Name`%in% indicators_keep]
gbr_clean <- gbr[`Indicator Name`%in% indicators_keep]

# Remove the duplicates
bra_clean <- unique(bra_clean)
ago_clean <- unique(ago_clean)
gbr_clean <- unique(gbr_clean)

```

# Merging Datasets

The following process is to merge both datasets for an easier look and comparative analysis of the indicators

```{r}
all_data <- rbindlist(list(bra_clean, ago_clean, gbr_clean), use.names = TRUE, fill = TRUE)
```

## Table preview of merged data with basic summary statistics

```{r}
# Average of each indicator value for each country
summary_stats <- all_data[
  , .(Average = mean(Value, na.rm = TRUE)), 
  keyby = .(`Country Name`, `Indicator Name`)
]

# Renaming the indicators for a cleaner look
summary_stats[,`Indicator Name`:= fcase(
  `Indicator Name` == "Life expectancy at birth, total (years)", "Life Expectancy",  `Indicator Name` == "Net migration", "Net migration",
  `Indicator Name` == "Unemployment, total (% of total labor force) (modeled ILO estimate)", "Unemployment"
)]

kable(summary_stats)
```

# Key Analysis and Plots

***Two*** analysis were done to assess the following:

1. How each indicator has evolved year by year within and across countries

2. A comparison in between countries for each indicator in the last year **(2024)**

---

## Analysis 1 - Trend over Time

```{r}
# Renaming indicators for better look
all_data[,`Indicator Name`:= fcase(
  `Indicator Name` == "Life expectancy at birth, total (years)", "Life Expectancy",  `Indicator Name` == "Net migration", "Net Migration",
  `Indicator Name` == "Unemployment, total (% of total labor force) (modeled ILO estimate)", "Unemployment"
)]

# Creating the trend analysis
trend <- all_data[
  `Indicator Name` %in% c("Life Expectancy", "Net Migration", "Unemployment"),
  .(Average = mean(Value, na.rm = TRUE)),
  by = .(`Year`, `Country Name`, `Indicator Name`)]

library(ggplot2)

ggplot(trend, aes(x = Year, y = Average, color = `Country Name`)) +
  geom_line(size = 1) +
  facet_wrap(~ `Indicator Name`, scales = "free_y") +
  labs(title = "Trends of Indicators Over Time by Country",
       y = "Value",
       x = "Year") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Analysis 2 - Cross-Country Comparison

For this analysis, only `Net Migration` and `Unemployment` are considered due to absence of data for 2024 on `Life Expectancy`

```{r}
latest_year <- max(all_data$Year, na.rm = TRUE)
comparison <- all_data[Year == latest_year,
  .(Value = mean(Value, na.rm = TRUE)),
  by = .(`Country Name`,`Indicator Name`)
]

ggplot(comparison, aes(x = `Country Name`, y = Value, fill = `Country Name`)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ `Indicator Name`, scales = "free_y") +
  labs(title = paste("Indicator Comparison Across Countries in", latest_year),
       y = "Value",
       x = "Country") +
  theme_minimal()

```

# Insights and Findings

## Trend over Time

This trend analysis have shown:

- Across 6 decades, `Life Expectancy` considerably increased for all 3 countries. Whilst [Brazil]{style="color: green;"} and the [United Kingdom]{style="color: blue;"} had a consistent exponential gain throughout, [Angola]{style="color: red;"} had its major boost in the past 20 years

- Regarding the `Net Migration`, each country had their own different development, with very unstable peaks throughout the years. The [United Kingdom]{style="color: blue;"} presented the highest increase overall, whilst [Brazil]{style="color: green;"} kept negative numbers for the vast majority of the years. [Angola]{style="color: red;"} on the other hand had its numbers quite stable, until it started to decrease around 2010, reaching a 0 mark in most recent years

- `Unemployment` data was only captured from the year 1991, showing [Angola]{style="color: red;"} with the highest incidence of unemployed people. The [United Kingdom]{style="color: blue;"} and [Brazil]{style="color: green;"} numbers varied, with many high and low peaks in different years. 

## Cross-Country Comparison

Giving a closer look into the indicators in the past year, the database for these 3 countries have demonstrated:

- The [United Kingdom]{style="color: blue;"} have had the *highest* `Net Migration`, but the *lowest* `Unemployment` rate 

- [Brazil]{style="color: green;"} was the country with *lowest* `Net Migration`, however with an *average* `Unemployment` rate

- [Angola]{style="color: red;"} barely presented `Net Migration` in 2024, with its rate *close to 0*, compensating in its `Unemployment` rate which was the *highest* among all countries.

# Conclusion

- Compiling both analysis made for this assignment, an overall difference in the demographics of all 3 countries is noted. The results exemplify the development of countries like the [United Kingdom]{style="color: blue;"}, the slow progress of [Brazil]{style="color: green;"} and the under development of an African nation like [Angola]{style="color: red;"}

- A special mention to the reflection of the **Covid-19 pandemic break in 2020**, where `Net Migration` differed in the [United Kingdom]{style="color: blue;"} (with a drastic decrease) and a big peak in [Brazil]{style="color: green;"}. Also, with a massive peak in Unemployment in all 3 countries.
